Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-03-16 TEBenchmarkPaper/article.tex
Analysis

We chose TIP for our ground truth since it satisfies the desiderata of
section~\ref{section:prep}: each problem has standalone type and function
definitions, making their separation trivial; known examples from the software
verification and inductive theorem proving literature are included, ensuring
relevance to those fields; the format includes the higher-order functions and
inductive datatypes we are interested in; it is large enough to pose a challenge
to current MTE tools; plus it is accompanied by tooling to convert its custom
format (an extension of SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of
languages, including Haskell and Isabelle.

@@

\begin{samepage}
\begin{verbatim}
(define-fun
  (par (a)                   ;; Type parameter
    (constructor-Cons        ;; Function name
      ((x a) (xs (List a)))  ;; Argument names and types
      (List a)               ;; Return type
      (as                    ;; Type annotation
        (Cons x xs)          ;; Return value
        (List a)))))         ;; Return type

(define-fun
  (par (a)                        ;; Type parameter
    (destructor-head              ;; Function name
      ((xs (List a)))             ;; Argument name and type
      a                           ;; Return type
      (match xs                   ;; Pattern-match
        (case (Cons h t) h)))))   ;; Return relevant field
\end{verbatim}
\end{samepage}

@@

We benchmarked QuickSpec version 1 (0.9.6), a tool written in Haskell for
conjecturing equations involving Haskell functions, described in more detail in
section~\ref{section:mte}. In order to thoroughly benchmark QuickSpec, we need
to automate some of the decisions which are normally left up to the user:

@@

The time taken to explore samples of different sizes is shown in
figure~\ref{figure:quickspec_runtimes}. Runs which timed out are shown in red,
and these occurred for all sample sizes but are more common for larger samples
(over half the runs for sample size 20 timed out). Runs which succeeded mostly
finished within 30 seconds, generating few or no conjectures; those taking
longer generated more conjectures, with the most being 116 conjectures from a
sample of size 16.

Precision and recall are shown in figure~\ref{figure:quickspec_precRec}, which
we model as simple Bernoulli processes: the generated conjectures (wanted or
not) are like coin tosses whose bias is the precision; likewise, the ground
truth theorems (found or not) are like coin tosses biased by the recall. Note
that this simplification does not explain how many conjectures the tool decides
to output for each run.

QuickSpec's precision gets worse as the sample size grows. Since QuickSpec
generates monotonically more conjectures as definitions are added to a signature
(assuming sufficient testing), it can't be finding fewer correct conjectures at
larger sizes. This is supported by the relative flatness of the recall results.
Rather, the number of conjectures generated is increasing at a higher rate than
the size of the ground truth. These extra conjectures may involve those
``padding'' definitions in a sample which don't contribute to its ground truth,
or may be ``uninteresting'' relationships between the dependencies of different
ground truth theorems.

@@

\begin{table}[H]
  \centering
  \begin{tabular}{ |r|l|l| }
    \hline
    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
    \hline
    \bfseries Found by QuickSpec     & 28 & 19 \\
    \bfseries Not found by QuickSpec & 4  & 28 \\
    \hline
  \end{tabular}
\end{table}

Precision was also compared by pooling results from samples where both tools
finished successfully. We cannot perform McNemar's test in this case, since each
tool generated different sets of conjectures. Instead, we add up the number of
results which are ``interesting'' (appear in the ground truth) and those which
are ``uninteresting'' (do not appear in the ground truth):

\begin{table}[H]
  \centering
  \begin{tabular}{ |r|l|l| }
    \hline
              & \bfseries Interesting & \bfseries Uninteresting \\
    \hline
    \bfseries IsaCoSy   & 32          & 137      \\
    \bfseries QuickSpec & 47          & 301      \\
    \hline
  \end{tabular}
\end{table}

These totals were tested for independence using Boschloo's
test~\cite{lydersen2009recommended}, assuming a binomial distribution and
conditioning on the number of generated conjectures (as if this were fixed by
design). This produces a p-value of 0.111, which given our (arbitrary) choice of
$\alpha = 0.05$ is not significant enough to reject the null hypothesis that
their precision is equal.

Note that this is the only comparison where IsaCoSy performed better (although
not significantly so) with 19\% precision compared to QuickSpec's 14\%. However,
precision on its own is not the whole story: it can be increased at the expense
of recall by making the tool more conservative. IsaCoSy's significantly lower
recall indicates that it may be overly-conservative compared to QuickSpec, which
would also explain its precision.

More importantly, the poor time and memory usage of IsaCoSy meant that very few
samples finished successfully. If we include failed runs in our tables, treating
them as if they succeeded with no output, the resulting statistics lean heavily
in favour of QuickSpec simply because it generated so much more than IsaCoSy in
the available time.

% To do this we can take a set of ATP tasks, for example proving the equation
% \ref{eq:mapreduce} from the theory in figure \ref{figure:list_theory}, and we
% split apart the definitions (to form a theory) from the theorem statements (to
% form our ground truth).

>>>>

We chose TIP for our ground truth since it satisfies the desiderata of
$\S$\ref{section:prep}: each problem has standalone type and function
definitions, making their separation trivial; known examples from the software
verification and inductive theorem proving literature are included, ensuring
relevance to those fields; the format includes the higher-order functions and
inductive datatypes we are interested in; it is large enough to pose a challenge
to current MTE tools; plus it is accompanied by tooling to convert its custom
format (an extension of SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of
languages, including Haskell and Isabelle.

@@

\begin{samepage}
\begin{verbatim}
(define-fun
  (par (a)                   ;; Type parameter
    (constructor-Cons        ;; Function name
      ((x a) (xs (List a)))  ;; Argument names and types
      (List a)               ;; Return type
      (as                    ;; Type annotation
        (Cons x xs)          ;; Return value
        (List a)))))         ;; Return type
\end{verbatim}
\end{samepage}

\begin{samepage}
\begin{verbatim}
(define-fun
  (par (a)                        ;; Type parameter
    (destructor-head              ;; Function name
      ((xs (List a)))             ;; Argument name and type
      a                           ;; Return type
      (match xs                   ;; Pattern-match
        (case (Cons h t) h)))))   ;; Return relevant field
\end{verbatim}
\end{samepage}

@@

We benchmarked QuickSpec version 1 (0.9.6), a tool written in Haskell for
conjecturing equations involving Haskell functions, described in more detail in
$\S$\ref{section:mte}. In order to thoroughly benchmark QuickSpec, we need
to automate some of the decisions which are normally left up to the user:

@@

The time taken to explore samples of different sizes is shown in
figure~\ref{figure:quickspec_runtimes}. Runs which timed out are shown in red,
and these occurred for all sample sizes but are more common for larger samples
(over half the runs for sample size 20 timed out). Runs which succeeded mostly
finished within 30 seconds, generating few or no conjectures; those taking
longer generated more conjectures, with the most being 133 conjectures from a
sample of size 19.

Precision and recall are shown in figure~\ref{figure:quickspec_precRec}.
Since QuickSpec generates monotonically more conjectures as definitions are
added to a signature (assuming sufficient testing), its decreasing precision
can't be due to finding fewer wanted conjectures at larger sizes. This is
supported by the relative flatness of the recall results. Rather, the number of
conjectures generated is increasing at a higher rate than the size of the ground
truth. These extra conjectures may involve those ``padding'' definitions in a
sample which don't contribute to its ground truth, or may be ``uninteresting''
relationships between the dependencies of different ground truth theorems.

@@

\begin{table}[H]
  \centering
  \begin{tabular}{ |r|l|l| }
    \hline
    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
    \hline
    \bfseries Found by QuickSpec     & 28 & 19 \\
    \bfseries Not found by QuickSpec & 4  & 28 \\
    \hline
    \hline
  \end{tabular}
\end{table}

McNemar's test is not applicable for comparing precision, since each tool
generated different sets of conjectures. Instead, we add up the number of
results which are ``interesting'' (appear in the ground truth) and those which
are ``uninteresting'' (do not appear in the ground truth):

\begin{table}[H]
  \centering
  \begin{tabular}{ |r|l|l| }
    \hline
              & \bfseries Interesting & \bfseries Uninteresting \\
    \hline
    \bfseries IsaCoSy   & 32          & 137      \\
    \hline
    \bfseries QuickSpec & 47          & 301      \\
    \hline
  \end{tabular}
\end{table}

We test these totals for independence using Boschloo's test and find a p-value
of 0.111, which exceeds our (arbitrary) choice of $\alpha = 0.05$; hence we do
not consider the difference between these proportions (14\% for QuickSpec, 19\%
for IsaCoSy) to be significant.

Note that precision is the only comparison where IsaCoSy scored higher, and even
that was not found to be significant. Also, precision on its own is not the
whole story: it can be increased at the expense of recall by making a tool more
conservative. IsaCoSy may already be overly-conservative compared to QuickSpec,
given that its recall is significantly lower.

More importantly, the poor time and memory usage of IsaCoSy meant that very few
samples finished successfully. If we include failed runs in our tables, treating
them as if they succeeded with no output, the resulting statistics lean
overwhelmingly in favour of QuickSpec simply because it generated so much more
than IsaCoSy in the available time.
